### 2016-03-08 paper/icfp16-submission/semantics.tex
misc semantics

Next, we formalize the notion of type error witnesses as follows.
%
First, we define a core calculus within which we will work~\S~\ref{sec:syntax}.
%
Second, we develop a (non-deterministic) operational semantics
for ill-typed programs that precisely defines the notion
of a \emph{witness}~\S~\ref{sec:semantics},
%
Third, we formalize and prove a notion of \emph{generality} for
witnesses, which states, intuitively that if we find a
single witness then for \emph{every possible} type
assignment, there exist inputs that are guaranteed to make
the program ``go wrong''~\S~\ref{sec:soundness}, and finally,
%
Fourth, we refine the operational semantics into a
\emph{search procedure} that returns concrete (general)
witnesses for ill-typed programs~\S~\ref{sec:search-algorithm}.

@@

\paragraph{Holes}
\label{sec:holes}
%
Recall that a key challenge in our setting is to find witnesses
that are meaningful and do not arise from choosing values from
irrelevant types.
%
We solve this problem be equipping our term language with a
notion of a \emph{hole}, written \vhole{\thole}, which represents
an \emph{unconstrained} value $\ehole$ that may be replaced with
\emph{any} value of type \thole.
%
Intuitively, the type holes \thole\ can be viewed as type variables
that we will \emph{not} generalize over.
%
A \emph{normalized} value (resp.\ type) is one that is not a hole,
but which may internally contain holes.
%
For example \RJ{add example here}

\paragraph{Substitutions}
%
Our semantics ensure the generality of witnesses by incrementally
\emph{refining} holes, filling in just as much information as is
needed locally to make progress (inspired by the manner in
which SmallCheck uses lazy evaluation~\cite{runciman_smallcheck_2008}).
%
We track how the holes are incrementally filled in, by using
value (resp.\ type) \emph{substitutions} $\vsu$ (resp. $\tsu$)
that map value (resp.\ type) holes to values (resp.\ types).
%
A \emph{normalized} substitution is one whose co-domain comprises
of normalized values (resp.\ types).
%
In the sequel, we will assume and ensure that all substitutions
are normalized.

@@

\paragraph{Generating Values} The (non-deterministic)
$\gen{t}{\tsu}$ in Figure~\ref{fig:gen}, takes
as input a type $t$ and returns a value of that type.
%
For base types, the procedure returns an arbitrary value of
that type.
%
For functions, it returns a lambda with a \emph{new} hole
denoting the return value, and unconstrained types (denoted
by $\thole$) yield fresh holes constrained to have type
\thole (denoted by $\vhole{\thole}$).
%
Note that for an algebraic datatype, $\gen{\ttree{t}}{\tsu}$
will always place holes in the generated nodes; they will be
lazily filled in later, on demand.

>>>>

Next, we formalize the notion of type error witnesses as follows.
%
First, we define a core calculus within which we will work~\S~\ref{sec:syntax}.
%
Second, we develop a (non-deterministic) operational semantics
for ill-typed programs that precisely defines the notion
of a \emph{witness}~\S~\ref{sec:semantics}.
%
Third, we formalize and prove a notion of \emph{generality} for
witnesses, which states, intuitively, that if we find a
single witness then for \emph{every possible} type
assignment there exist inputs that are guaranteed to make
the program ``go wrong''~\S~\ref{sec:soundness}.
%
Finally, we refine the operational semantics into a
\emph{search procedure} that returns concrete (general)
witnesses for ill-typed programs~\S~\ref{sec:search-algorithm}.

@@

\paragraph{Holes}
\label{sec:holes}
%
Recall that a key challenge in our setting is to find witnesses
that are meaningful and do not arise from choosing values from
irrelevant types.
%
We solve this problem be equipping our term language with a
notion of a \emph{hole}, written \vhole{\thole}, which represents
an \emph{unconstrained} value $\ehole$ that may be replaced with
\emph{any} value of type \thole.
%
Intuitively, the type holes \thole\ can be viewed as type variables
that we will \emph{not} generalize over.
%
A \emph{normalized} value (resp.\ type) is one that is not a hole,
but which may internally contain holes.
%
For example
$\vnode{\thole}{\vhole{\thole}}{\vleaf{\thole}}{\vleaf{\thole}}$ is a
normalized value, and its type $\ttree{\thole}$ is a normalized type.

\paragraph{Substitutions}
%
Our semantics ensure the generality of witnesses by incrementally
\emph{refining} holes, filling in just as much information as is
needed locally to make progress (inspired by the manner in
which SmallCheck uses lazy evaluation~\cite{runciman_smallcheck_2008}).
%
We track how the holes are incrementally filled in, by using
value (resp.\ type) \emph{substitutions} $\vsu$ (resp. $\tsu$)
that map value (resp.\ type) holes to values (resp.\ types).
%
A \emph{normalized} substitution is one whose co-domain is comprised
of normalized values (resp.\ types).
%
In the sequel, we will assume and ensure that all substitutions
are normalized.

@@

\paragraph{Generating Values} The (non-deterministic)
$\gen{t}{\tsu}$ in Figure~\ref{fig:gen}, takes
as input a type $t$ and returns a value of that type.
%
For base types, the procedure returns an arbitrary value of
that type.
%
For functions, it returns a lambda with a \emph{new} hole
denoting the return value, and unconstrained types (denoted
by $\thole$) yield fresh holes constrained to have type
\thole (denoted by $\vhole{\thole}$).
%
When generating a $\ttree{t}$ we must take care to ensure
the resulting tree is well-typed. For a polymorphic type
$\ttree{\thole}$, eg in \recasegoodone, we will
place holes in the generated nodes; they will be lazily
filled in later, on demand.
